---
title: "COVID19 data exploration"
author: "Vincent L. Cannataro"
date: 'last update `r format(Sys.Date(), "%Y-%B-%d")`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


Data obtained from curated Johns Hopkins University Center for Systems Science and Engineering database here: https://github.com/CSSEGISandData/COVID-19.git and from from The New York Times, based on reports from state and local health agencies https://github.com/nytimes/covid-19-data 


```{r get data, include=F}
# jhu_data <- read.csv(check.names = F,
#   file = "JHU_COVID19_data/COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv")
# 
# # clean data types prior to melting
# jhu_data <- jhu_data %>%
#   mutate_if(is.numeric,as.character, 
#             is.factor, as.character)
# 
# jhu_data_m <- tidyr::pivot_longer(data = jhu_data, 
#                                   cols=`1/22/20`:colnames(jhu_data)[ncol(jhu_data)], 
#                                   values_to = "Confirmed_cases",names_to = "Date") %>%
#   mutate(Confirmed_cases = as.numeric(Confirmed_cases))
# 
# jhu_data_m <- jhu_data_m %>%
#   mutate(Date = as.Date(Date, format= "%m/%d/%y"))

# March 22 update 
# run cleaning_data_for_export.R first
# load("output_data/cleaned_data/combined_CDR_data.RData")



# try with NY Times data 
nytimes_county <- read.csv(file = "NY_Times_COVID19_data/covid-19-data/us-counties.csv",
                           stringsAsFactors = T)


nytimes_county$date <- as.Date(nytimes_county$date,format = "%Y-%m-%d")

```

# Current USA confirmed cases


```{r USA data, include=F}
# 
# # 
# US_data <- combined_CDR_data %>% 
#   filter(Country %in% c("US")) %>%
#   filter(State %in% state.name) %>%
#   # group_by(State,Updated) %>%
#   # summarize(all_cases = sum(Confirmed)) %>%
#   mutate(Date = as.Date(format(Updated, "%Y-%m-%d"))) %>%
#   # group_by(State,Date) %>%
#   # summarize(all_cases = max(Confirmed)) %>%
#   group_by(Date) %>%
#   summarize(all_cases = sum(Confirmed)) %>%
#   filter(!is.na(all_cases)) 
# 
# US_data$cumulative_sum <- NA
# US_data$cumulative_sum[1] <- US_data$all_cases[1]
# for(day_ind in 2:nrow(US_data)){
#  US_data$cumulative_sum[day_ind] <- US_data$cumulative_sum[day_ind-1] + US_data$all_cases[day_ind]
# }

US_data <- nytimes_county %>%
  group_by(date) %>%
  summarize(all_cases = sum(cases)) 




```




The total number of cases in the USA is rapidly increasing. 

```{r,echo=F}
ggplot(data = US_data) + 
  geom_point(aes(x=date, y=all_cases)) + 
  theme_bw()  + 
  labs(y="All confirmed cases") + 
  labs(title="Confirmed cases in the USA",
       caption = "Data: The New York Times, https://github.com/nytimes/covid-19-data\nPlot: @VinCannataro") 

ggplot(data = US_data) + 
  geom_point(aes(x=date, y=all_cases)) + 
  theme_bw()  + 
  labs(y="All confirmed cases, log10 scale") + 
  labs(title="Confirmed cases in the USA, Y axis log10 scaled",caption = "Data: The New York Times, https://github.com/nytimes/covid-19-data\nPlot: @VinCannataro") + 
  scale_y_log10()
```

```{r, include=F}

start_date <- as.Date(x = "03/01/2020",format= "%m/%d/%y")

end_date <- as.Date(x = "04/12/2020",format= "%m/%d/%y")

```




# By State

```{r breaking up USA by state, include=F}
# just the USA data 
# USA_by_state <- jhu_data_m %>%
#   filter(`Country/Region` == "US")
# 
# 
# # need to collapse county level to the state level
# state_names <- data.frame(state_name = state.name, state_abb = state.abb,stringsAsFactors = F)
# state_names <- rbind(state_names,c("D.C.","D.C."))
# rownames(state_names) <- state_names$state_abb
# 
# 
# # grep(pattern = paste(state.abb,collapse = "|"),x = USA_by_state$`Province/State`,)
# 
# state_vec <- rep(NA,length=nrow(USA_by_state))
# 
# state_split <- strsplit(x = as.character(USA_by_state$`Province/State`),split = ",")
# 
# # loop through to bring county --> state level 
# for(state_ind in 1:length(state_split)){
#  if(length(state_split[[state_ind]])==1){
#    state_vec[state_ind] <- state_split[[state_ind]]
#  }else{
#    state_vec[state_ind] <- state_names[trimws(state_split[[state_ind]][2]),"state_name"]
#  }
# }
# 
# 
# USA_by_state$state <- state_vec 
# 
# USA_by_state <- USA_by_state %>%
#   filter(state %in% state_names$state_name)
# 
# USA_by_state$state <- factor(USA_by_state$state, 
#                                         levels=sort(unique(USA_by_state$state)))
# 
# USA_by_state <- USA_by_state %>%
#   group_by(state, Date) %>%
#   summarize(Confirmed_cases = sum(Confirmed_cases))
# ggplot(data = USA_by_state, aes(x = Date,y=Confirmed_cases,color=`Province/State`)) + 
#   geom_line() 

# this seems like what we want. 

USA_by_state <- nytimes_county %>%
  group_by(state,date) %>%
  summarize(Confirmed_cases = sum(cases)) %>%
  filter(state %in% state.name)

USA_by_state$state <- factor(USA_by_state$state, levels = unique(USA_by_state$state))
```



```{r resetting data to start at first case, include=F}

first_cases <- USA_by_state %>% 
  filter(Confirmed_cases > 0) %>%
  group_by(state) %>% 
  summarize(first_case = min(date)) %>%
  as.data.frame(., stringsAsFacter=F)
rownames(first_cases) <- first_cases$state

USA_by_state$Date_since_first_case <- USA_by_state$date - first_cases[as.character(USA_by_state$state), "first_case"]

USA_by_state <- USA_by_state %>%
  filter(Date_since_first_case > 0)

```




```{r plotting states,echo=F,warning=F,fig.height=7,fig.width=12}

ggplot(data = USA_by_state, aes(x=as.numeric(Date_since_first_case), 
                                y= Confirmed_cases, 
                                color=state,
                                shape=state)) + 
  geom_line(alpha=0.5) + 
  geom_point(alpha=0.5,size=3) + 
  theme_bw() + 
  scale_color_discrete(name="State") + 
  scale_shape_manual(name="State",
                       labels=levels(USA_by_state$state),
                       values=c(rep(15:19,11),15,16)) + 
  labs(title="Confirmed cases in the USA, split by state",
       caption = "Data: The New York Times, https://github.com/nytimes/covid-19-data\nPlot: @VinCannataro",
       y="Confirmed cases", 
       x="Days since first case in dataset ") + 
  # scale_y_log10() +
  scale_x_continuous(breaks = seq(from = 0,to = max(as.numeric(USA_by_state$Date_since_first_case)),by = 2))




ggplot(data = USA_by_state, aes(x=as.numeric(Date_since_first_case), 
                                y= Confirmed_cases, 
                                color=state,
                                shape=state)) + 
  geom_line(alpha=0.5) + 
  geom_point(alpha=0.5,size=3) + 
  theme_bw() + 
  scale_color_discrete(name="State") + 
  scale_shape_manual(name="State",
                       labels=levels(USA_by_state$state),
                       values=c(rep(15:19,11),15,16)) + 
  labs(title="Confirmed cases in the USA, split by state",
       caption = "Data: The New York Times, https://github.com/nytimes/covid-19-data\nPlot: @VinCannataro",
       y="Confirmed cases, log10 axis scale", 
       x="Days since first case in dataset ") + 
  scale_y_log10() +
  scale_x_continuous(breaks = seq(from = 0,to = max(as.numeric(USA_by_state$Date_since_first_case)),by = 2))




```





## Table of confirmed cases in the USA: 
The total number of confirmed cases in the USA by date: 

```{r, echo=F}
knitr::kable(US_data)
```

